Stochastic Variance Reduced Policy Gradient

### 2018-02-01 svrpg.tex
Updated

%%%CUSTOM COMMANDS%%%
%Math
\newcommand{\realspace}{\mathbb R}      % realspace
\newcommand{\transpose}[1]{{#1}^\texttt{T}}
\DeclareMathOperator*{\argmax}{arg\,max}
\DeclareMathOperator*{\argmin}{arg\,min}
\DeclareMathOperator*{\EV}{\mathbb{E}}
\DeclareMathOperator{\Tr}{Tr}
\DeclareMathOperator*{\Cov}{\mathbb{C}ov}
\DeclareMathOperator*{\Var}{\mathbb{V}ar}
\newcommand{\EVV}[2][\ppvect \in \ppspace]{\EV_{#1}\left[{#2}\right]}
\newcommand{\norm}[2][\infty]{\left\|#2\right\|_{#1}}
\newcommand{\Dij}[2]{\frac{\partial^{2}{#1}}{\partial{#2}_i\partial{#2}_j}}
\newcommand{\de}{\,\mathrm{d}}
\newcommand{\dotprod}[2]{\left\langle#1,#2\right\rangle}
%RL
\newcommand{\vtheta}{\boldsymbol{\theta}}
\newcommand{\Aspace}{\mathcal{A}}
\newcommand{\Sspace}{\mathcal{S}}
\newcommand{\Tspace}{\mathcal{T}}
\newcommand{\Transition}{\mathcal{P}}
\newcommand{\Reward}{\mathcal{R}}
\newcommand{\stationary}{d_{\rho}^{\pi_{\vtheta}}(s)}
\newcommand{\policy}{\pi_{\vtheta}(a \mid s)}
\newcommand{\pol}{\pi_{\vtheta}}
\newcommand{\trajdistr}{\pi_{\vtheta}(\tau)}
\newcommand{\score}[2]{\nabla\log\pi_{#1}(#2)}
\newcommand{\Qfun}{Q^{\pi_{\vtheta}}(s,a)}
\newcommand{\Vfun}{V^{\pi_{\vtheta}}(s)}
\newcommand{\vTheta}{\boldsymbol{\Theta}}
\newcommand{\vphi}{\boldsymbol{\phi}}
\newcommand{\gradJ}[1]{\nabla J(#1)}
\newcommand{\gradApp}[2]{\hat{\nabla}_{#2}J(#1)}
\newcommand{\eqdef}{\mathrel{\mathop:}=}
\newcommand{\Dataset}{\mathcal{D}}
%Specific
\newcommand{\Ets}[2][t]{\mathbb{E}_{#1\mid s}\left[#2\right]}
\newcommand{\Covts}[3][t]{{\mathbb{C}\text{ov}}_{#1\mid s}\left(#2,#3\right)}
\newcommand{\Varts}[2][t]{{\mathbb{V}\text{ar}}_{#1\mid s}\left[#2\right]}
\newcommand{\gradBlack}[1]{\blacktriangledown J(#1)}
\newcommand{\gradIdeal}[1]{\overline{\blacktriangledown} J(#1)}
\newcommand{\VARRF}{V}
\newcommand{\GRADLOG}{G}
\newcommand{\VARIS}{W}
\newcommand{\HESSLOG}{F}
%%%%%%

@@

\section{Introduction}
%Similarities between RL and SL
On a very general level, machine learning addresses the problem of artificial agents that must select the right actions to solve a given task. The approach of Reinforcement Learning \cite{sutton2011reinforcement} is to learn the best actions by direct interaction with the environment and evaluation of the performance in the form of a reward signal. This makes RL fundamentally different from Supervised Learning (SL), where correct actions are explicitly prescribed by a human teacher (\eg for classification, in the form of class labels). However, the two approaches share many challenges and, as a consequence, solution concepts. The problem of estimating a model from samples, which is at the core of SL, is equally fundamental in RL, whether we choose to model the environment, a value function, or a policy of behavior directly. In particular, when the tasks are characterized by large or continuous state-action spaces, RL needs to borrow the powerful function approximators that are the main subject of study of SL, such as neural networks. Among RL solutions for learning continuous tasks, policy gradient methods bear the closest similarity to SL, since they allow to formulate the RL problem as the unconstrained optimization of an objective function (\ie a negative loss) of policy parameters and data samples.

% SL needs SVRG so RL
In supervised learning, stochastic optimization is often employed for reasons of efficiency. The idea is to evaluate the objective on a random subset of the available data to obtain an approximate gradient direction. This can be much less expensive that taking into account all the data for all parameter updates. However, the sub-sampling of data samples is a source of variance that can potentially compromise convergence, so that speed and variance must be traded off with proper handling of meta-parameters.
Variance-reduced gradient algorithms such as SAG, SVRG and SAGA offer better ways of achieving the optimal trade-off between variance and speed, both in theory and practice. 
In Reinforcement Learning, stochastic optimization is not even a choice, since samples must be sequentially obtained by interacting with an initially unknown environment. In this scenario, containing the variance of the estimates is a problem that cannot be avoided, and methods like SVRG can offer a more efficient alternative

\todomat{work in progress}
% Potential advantages of SVRG
% Difficulties in RL \wrt SL

\section{Preliminaries}
In this section we provide essential background on policy gradient methods for Reinforcement Learning and stochastic variance-reduced gradient methods for finite-sum optimization. In the next section we will show how these methods can be combined to solve Reinforcement Learning tasks.

\subsection{Policy Gradient}\label{subsec:PolicyGradient}
A Reinforcement Learning task \cite{sutton2011reinforcement} can be modeled with a discrete-time continuous Markov Decision Process (MDP) $\langle\Sspace,\Aspace,\Transition,\Reward,\gamma,\rho\rangle$, where $\Sspace$ is a continuous state space; $\Aspace$ is a continuous action space; $\Transition$ is a Markovian transition model, where $\Transition(s'|s,a)$ defines the transition density from state $s$ to $s'$ under action $a$; $\Reward$ is the reward function, where $\Reward(s,a) \in [-R,R]$ is the expected reward for state-action pair $(s,a)$ and $R$ is the maximum absolute-value reward; $\gamma\in[0,1)$ is the discount factor; and $\rho$ is the initial state distribution.
The agent's behavior is modeled as a policy $\pi$, where $\pi(\cdot|s)$ is the density distribution over $\Aspace$ in state $s$.
% We consider episodic tasks, \ie tasks composed of episodes of length $H$, also called time horizon.
% In this setting, a trajectory $\tau$ is the sequence of states and actions $(s_0,a_0,s_1,a_1,\dots,s_{H-1},a_{H-1})$ observed by following some stationary policy in a given episode, where $s_0$ is always sampled from the initial state distribution $\rho$.
We consider episodic MDPs with effective horizon $H$, \ie optimal policy is able to reach the target state (\ie absorbing state) in at most $H$ steps (the true length depend on the initial state). This has not to be confused with a finite horizon problem where the optimal policy is non-stationary. In this setting, we can limit our attention to trajectories of length $H$. A trajectory $\tau$ is a sequence of states and actions $(s_0,a_0,s_1,a_1,\dots,s_{H-1},a_{H-1})$ observed by following a stationary policy, where $s_0 \sim \rho$.
With abuse of notation, we denote with $\pi(\tau)$ the density distribution induced by policy $\pi$ on the set $\Tspace$ of all possible trajectories, and with $\Reward(\tau)$ the total discounted reward provided by trajectory $\tau$:
\begin{align*}
\Reward(\tau) = \sum_{t=0}^{H-1}\gamma^t\Reward(s_t,a_t).
\end{align*}
Policies can be ranked based on their expected total reward:
\begin{align*}
	J(\pi) = \int_{\Tspace}\pi(\tau)\Reward(\tau)\de \tau.
\end{align*}
Solving an MDP means finding $\pi^* \in \argmax_{\pi} \{J(\pi)\}$.

Policy gradient methods restrict the search for the performance-maximizing policy over a class of parametrized policies $\Pi_{\vtheta}=\{\pol: \vtheta \in \realspace^m\}$, with the only constraint that $\pol$ is differentiable \wrt the $m$-dimensional parameter vector $\vtheta$. We denote the performance of a parametric policy with $J(\vtheta)$ for brevity, while in some occasions we replace $\pol(\tau)$ with $\pi(\tau|\vtheta)$ for the sake of readability. The search for a locally optimal policy is performed by means of gradient ascent, where the gradient of $J(\vtheta)$ \wrt the policy parameters is \cite{sutton2000policy}:
\todopir{stochastic or mini-batch gradient}
\begin{align*}
\gradJ{\vtheta} = \int_{\Tspace}\pol(\tau)\score{\vtheta}{\tau}\Reward(\tau)\de \tau.
\end{align*}
We consider the on-line learning problem, where trajectories must be sampled directly by interacting with the environment. 
In this setting, $(N>0)$-batch gradient ascent is typically employed: after sampling a batch of $N$ trajectories (\ie $\Dataset_{N}$),
policy parameters are updated as $\vtheta\gets\vtheta + \alpha\gradApp{\vtheta}{N}$, where $\alpha$ is a step size and $\gradApp{\vtheta}{N}$ is an estimate of the true gradient. 
\todopir{step size is scalar or vectorial (just for notation)? Check consistency of $N$ with finite-sum function, we can use even $b$ or $B$ for the batch size.}
The REINFORCE gradient estimator~\citep{williams1992simple} provides a simple, unbiased way of estimating the gradient:
\begin{align*}
\gradApp{\vtheta}{N} = \frac{1}{N}\sum_{n=1}^{N}\left(\sum_{h=0}^{H-1}\score{\vtheta}{a_h^n\mid s_h^n}\right)\left(\sum_{h=0}^{H-1}\gamma^h r_h^n - b\right),
\end{align*}
where subscripts denote the time step, superscripts denote the trajectory, $r_h$ is the reward actually collected at time $h$ and b can be any baseline, provided it is constant \wrt to actions.
\todopir{Actually it can be state-action dependant~\citep[\eg][]{Thomas2017actionbaseline,wu2018variance}.}
The G(PO)MDP gradient estimator~\cite{baxter2001infinite} is a refinement of REINFORCE which is subject to less variance while preserving the unbiasedness:
\begin{align*}
\gradApp{\vtheta}{N} = \frac{1}{N}\sum_{n=1}^{N}\sum_{h=0}^{H-1}\left(\sum_{k=0}^{h}\score{\vtheta}{a_k^n\mid s_k^n}\right)\left(\gamma^h r_h^n - b\right).
\end{align*}
\todopir{Cite Jan's paper somewhere}
\todopir{At this point we need to connect with the supervised learning framework (we can just define everything inside the first sum as $\nabla g(\tau)$).}

>>>>

%%%CUSTOM COMMANDS%%%
%Math
\newcommand{\realspace}{\mathbb R}      % realspace
\newcommand{\transpose}[1]{{#1}^\texttt{T}}
\DeclareMathOperator*{\argmax}{arg\,max}
\DeclareMathOperator*{\argmin}{arg\,min}
\DeclareMathOperator*{\EV}{\mathbb{E}}
\DeclareMathOperator{\Tr}{Tr}
\DeclareMathOperator*{\Cov}{\mathbb{C}ov}
\DeclareMathOperator*{\Var}{\mathbb{V}ar}
\newcommand{\EVV}[2][\ppvect \in \ppspace]{\EV_{#1}\left[{#2}\right]}
\newcommand{\norm}[2][\infty]{\left\|#2\right\|_{#1}}
\newcommand{\Dij}[2]{\frac{\partial^{2}{#1}}{\partial{#2}_i\partial{#2}_j}}
\newcommand{\de}{\,\mathrm{d}}
\newcommand{\dotprod}[2]{\left\langle#1,#2\right\rangle}
%RL
\newcommand{\vtheta}{\boldsymbol{\theta}}
\newcommand{\Aspace}{\mathcal{A}}
\newcommand{\Sspace}{\mathcal{S}}
\newcommand{\Tspace}{\mathcal{T}}
\newcommand{\Transition}{\mathcal{P}}
\newcommand{\Reward}{\mathcal{R}}
\newcommand{\stationary}{d_{\rho}^{\pi_{\vtheta}}(s)}
\newcommand{\policy}{\pi_{\vtheta}(a \mid s)}
\newcommand{\pol}{\pi_{\vtheta}}
\newcommand{\trajdistr}{\pi_{\vtheta}(\tau)}
\newcommand{\score}[2]{\nabla\log\pi_{#1}(#2)}
\newcommand{\Qfun}{Q^{\pi_{\vtheta}}(s,a)}
\newcommand{\Vfun}{V^{\pi_{\vtheta}}(s)}
\newcommand{\vTheta}{\boldsymbol{\Theta}}
\newcommand{\vphi}{\boldsymbol{\phi}}
\newcommand{\gradJ}[1]{\nabla J(#1)}
\newcommand{\gradApp}[2]{\widehat{\nabla}_{#2}J(#1)}
\newcommand{\eqdef}{\mathrel{\mathop:}=}
\newcommand{\Dataset}{\mathcal{D}}
%Specific
\newcommand{\Ets}[2][t]{\mathbb{E}_{#1\mid s}\left[#2\right]}
\newcommand{\Covts}[3][t]{{\mathbb{C}\text{ov}}_{#1\mid s}\left(#2,#3\right)}
\newcommand{\Varts}[2][t]{{\mathbb{V}\text{ar}}_{#1\mid s}\left[#2\right]}
\newcommand{\gradBlack}[1]{\blacktriangledown J(#1)}
\newcommand{\gradIdeal}[1]{\overline{\blacktriangledown} J(#1)}
\newcommand{\VARRF}{V}
\newcommand{\GRADLOG}{G}
\newcommand{\VARIS}{W}
\newcommand{\HESSLOG}{F}
%%%%%%

@@

\section{Introduction}
\begin{itemize}
        \item Similarities between RL and SL
        \item SL needs SVRG so RL
        \item Potential advantages of SVRG
        \item Difficulties in RL \wrt SL
\end{itemize}

\section{Preliminaries}
In this section we provide the essential background on policy gradient methods for Reinforcement Learning and stochastic variance-reduced gradient methods for finite-sum optimization.
%In the next section we will show how these methods can be combined to solve Reinforcement Learning tasks.

\subsection{Policy Gradient}\label{subsec:PolicyGradient}
A Reinforcement Learning task \cite{sutton2011reinforcement} can be modeled with a discrete-time continuous Markov Decision Process (MDP) $M = \{\Sspace,\Aspace,\Transition,\Reward,\gamma,\rho\}$, where $\Sspace$ is a continuous state space; $\Aspace$ is a continuous action space; $\Transition$ is a Markovian transition model, where $\Transition(s'|s,a)$ defines the transition density from state $s$ to $s'$ under action $a$; $\Reward$ is the reward function, where $\Reward(s,a) \in [-R,R]$ is the expected reward for state-action pair $(s,a)$;
% and $R$ is the maximum absolute-value reward;
$\gamma\in[0,1)$ is the discount factor; and $\rho$ is the initial state distribution.
The agent's behaviour is modelled as a policy $\pi$, where $\pi(\cdot|s)$ is the density distribution over $\Aspace$ in state $s$.
% We consider episodic tasks, \ie tasks composed of episodes of length $H$, also called time horizon.
% In this setting, a trajectory $\tau$ is the sequence of states and actions $(s_0,a_0,s_1,a_1,\dots,s_{H-1},a_{H-1})$ observed by following some stationary policy in a given episode, where $s_0$ is always sampled from the initial state distribution $\rho$.
We consider episodic MDPs with effective horizon $H$.\footnote{The episode duration is a random variable but the optimal policy is able to reach the target state (\ie absorbing state) in at most $H$ steps. This has not to be confused with a finite horizon problem where the optimal policy is non-stationary.} In this setting, we can limit our attention to trajectories of length $H$. A trajectory $\tau$ is a sequence of states and actions $(s_0,a_0,s_1,a_1,\dots,s_{H-1},a_{H-1})$ observed by following a stationary policy, where $s_0 \sim \rho$.
With abuse of notation, we denote with $\pi(\tau)$ the density distribution induced by policy $\pi$ on the set $\Tspace$ of all possible trajectories, and with $\Reward(\tau)$ the total discounted reward provided by trajectory $\tau$:
%
$\Reward(\tau) = \sum_{t=0}^{H-1}\gamma^t\Reward(s_t,a_t).$
%
Policies can be ranked based on their expected total reward: $J(\pi) = \EVV[\tau \sim \pi]{\Reward(\tau)|M}$.
Solving an MDP $M$ means finding $\pi^* \in \argmax_{\pi} \{J(\pi)\}$.

Policy gradient methods restrict the search for the performance-maximizing policy over a class of parametrized policies $\Pi_{\vtheta}=\{\pol: \vtheta \in \realspace^m\}$, with the only constraint that $\pol$ is differentiable \wrt $\vtheta$. We denote the performance of a parametric policy with $J(\vtheta)$ for brevity, while in some occasions we replace $\pol(\tau)$ with $\pi(\tau|\vtheta)$ for the sake of readability.\todopirout{See if it used}
The search for a locally optimal policy is performed by means of gradient ascent, where the policy gradient
%of $J(\vtheta)$ \wrt the policy parameters 
is \cite{sutton2000policy, Peters2008reinf}:
\begin{align} \label{E:policygradient}
        \gradJ{\vtheta} = \EVV[\tau \sim \pol]{\score{\vtheta}{\tau}\Reward(\tau)}
        %\int_{\Tspace}\pol(\tau)\score{\vtheta}{\tau}\Reward(\tau)\de \tau.
\end{align}
Notice that the distribution defining the gradient is induced by the current policy. This aspect introduces a nonstationarity in the sampling process. Since the underlying distribution changes over time, it is necessary to resample at each update or use weighting techniques such as importance sampling.
Here, we consider the \emph{on-line learning scenario}, where trajectories are sampled by interacting with the environment at each policy change. 
In this setting, stochastic gradient ascent is typically employed.
At each iteration $k >0$, a batch $\mathcal{D}_k = \{\tau^i\}_{i=0}^N$ of $N>0$ trajectories is collected using policy $\pi_{\vtheta_k}$.
The policy is then updated as $\vtheta_{k+1}  = \vtheta_k + \alpha\gradApp{\vtheta_k}{N}$, where $\alpha$ is a step size and $\gradApp{\vtheta}{N}$ is an estimate of~\eqref{E:policygradient} using $\mathcal{D}_k$, \ie
\begin{align} \label{E:policygradient.estimate}
        \gradApp{\vtheta}{N} = \frac{1}{N}\sum_{n=1}^{N} \nabla f(\tau^i).
\end{align}
The most widespread gradient estimators are REINFORCE~\citep{williams1992simple} and G(PO)MDP~\citep{baxter2001infinite}.
We refer the reader to App.~\ref{A:gradient_estimators} for a formal definition.
Both are unbiased estimate of~\eqref{E:policygradient} but G(PO)MDP is subject to less variance than REINFORCE.
\todopir{At this point we need to connect with the supervised learning framework (we can just define everything inside the first sum as $\nabla g(\tau)$).}
